Structure-Preserving Model-Reduction of Dissipative Hamiltonian Systems

### 2017-03-19 dissipative_hamiltonian/chapters/1.intro.tex
first draft finished

The need for more accuracy has led to ever more complex models in engineering and physics. As a consequence, direct numerical methods for solving partial differential equations, a powerful modeling tool, have become inefficient due to limitations in computational power and resources.

Over the past decade, reduced basis methods have emerged as a popular approach to reduce the complexity of large systems of partial differential equations \cite{Ito:1998up,Ito:1998ch,Ito:2001ev}. These methods construct a low dimensional subspace, the reduced space, that approximately represent the solution to the system of differential equations. The projection of the original system onto the reduced space then allows the exploration of the solution with a significantly reduced computational complexity \cite{hesthaven2015certified,quarteroni2015reduced}.

Hamiltonian systems are an important class of systems that appear in engineering and physics. In these systems, preserving system energy is essential in a correct numerical solutions. Therefore, developing model reduction techniques that preserves the symplectic symmetry is essential. The classical reduced basis methods, do not generally preserve the conservation laws and symmetries, e.g. the system energy, of the full system \cite{Amsallem:2014ef,prajna2003pod}. This often results in an unstable or qualitatively wrong solution.

It is demonstrated by \cite{Lall:2003iy,Carlberg:2014ky,Peng:2014di}, that if the basis to the reduced space is not chosen carefully, the symplastic symmetry of Hamiltonian and Lagrangian systems will be destroyed over the course of model reduction. To resolve this issue Peng et al. \cite{Peng:2014di} and Lall et al. \cite{Lall:2003iy} construct a reduced ordered configuration space that inherits symmetries of the full configuration space. Then, by using a proper time integrator scheme one can preserve symmetries.

@@

In the context of network modeling and circuit simulation, a considerable attention is geared toward developing structure preserving, and in particular energy preserving, model reduction techniques. Model reduction of port-Hamiltonian system is given \cite{Polyuga:2010gj,beattie2011structure,chaturantabut2016structure} and the references therein. These methods construct use a Krylov or a Proper Orthogonal Decomposition (POD) method to construct a reduced port-Hamiltonian system that preserves the passivity, and thus the stability of the original system. In general these methods do not guarantee the correct distribution of the energy among the energy consuming and energy storing units. Further, over long time integration, accumulation of local errors might produce an erroneous solution.

In this paper we propose the Reduced Dissipative Hamiltonian Method (RDH), a structure preserving model reduction approach for dissipative Hamiltonian systems. By considering a canonical heat bath, also known as the hidden strings \cite{Figotin:2006jy,Figotin:2005}, the reduced system is extended to a closed and conservative system. Therefore, a symplectic time integrator can guarantee the conservation of the system energy and the dissipated energy. Further, the hidden strings assure that the local errors in the dissipation of energy do not accumulate, resulting in a correct evolution of the system energy.

This paper is organized as follow. Section \ref{sec:2} covers the required topics on Hamiltonian systems, dissipative Hamiltonian system and Hamiltonian extension. In Section \ref{sec:3} we discussion the symplectic model reduction for Hamiltonian systems and introduce the Reduced Dissipative Hamiltonian Method. Accuracy, stability, and efficiency of the RDH method is then discussed in Section \ref{sec:4}, through simulation of the dissipative wave equation and a linear port-Hamiltonian system of an electrical circuit. We offer conclusive remarks in Section \ref{sec:5}.

>>>>

The need for more accuracy has led to ever more complex models and systems of partial differential equations in engineering and physics. As a consequence, direct numerical methods for solving PDEs have become inefficient due to limitations in computational power and resources.

Over the past decade, reduced basis methods have emerged as a popular approach to reduce the complexity of large systems of partial differential equations \cite{Ito:1998up,Ito:1998ch,Ito:2001ev}. These methods construct a low-dimensional subspace, the reduced space, that approximately represent the solution to the system of differential equations. The projection of the original system onto the reduced space then allows the exploration of the solution with a significantly reduced computational complexity \cite{hesthaven2015certified,quarteroni2015reduced}.

Hamiltonian systems are an important class of systems that appear in engineering and physics. In these systems, preserving system energy is essential in a correct numerical solution. Therefore, developing model reduction techniques that preserve the symplectic symmetry is essential. The classical reduced basis methods, do not generally preserve the conservation laws and intrinsic symmetries of Hamiltonian systems \cite{Amsallem:2014ef,prajna2003pod}. This often results in an unstable or a qualitatively wrong solution.

It is demonstrated by \cite{Lall:2003iy,Carlberg:2014ky,Peng:2014di}, that if the basis for the reduced space is not chosen carefully, the symplastic symmetry of Hamiltonian and Lagrangian systems will be destroyed over the course of model reduction. To resolve this issue, Peng et al. \cite{Peng:2014di} and Lall et al. \cite{Lall:2003iy} construct a reduced ordered configuration space that inherits symmetries of the full configuration space. Then, by using a proper time integrator scheme, the symmetries are preserved in the reduced system.

@@

In the context of network modeling and circuit simulation, a considerable work has been done on developing structure preserving, and in particular energy preserving, model reduction techniques. Model reduction for port-Hamiltonian systems are given in \cite{Polyuga:2010gj,beattie2011structure,chaturantabut2016structure} and the references therein. These methods use a Krylov or a Proper Orthogonal Decomposition (POD) approach to construct a reduced port-Hamiltonian system that preserves the passivity, and thus the stability of the original system. In general, these methods do not guarantee the correct distribution of the energy among the energy consuming and energy storing units. Further, over long time integration, accumulation of local errors might produce an erroneous solution.

In this paper, we present the Reduced Dissipative Hamiltonian (RDH) method, a structure preserving model reduction approach for dissipative Hamiltonian systems. By considering a canonical heat bath, also known as the hidden strings \cite{Figotin:2006jy,Figotin:2005}, the reduced system is extended to a closed and conservative system. Therefore, a symplectic time integrator can guarantee the conservation of the system energy and the correct dissipation of energy. Further, the hidden strings assure that the local errors in the dissipation of energy do not accumulate, resulting in a correct evolution of the system energy.

This paper is organized as follow. Section \ref{sec:2} covers the required topics on Hamiltonian systems, dissipative Hamiltonian systems, and the Hamiltonian extension. In Section \ref{sec:3} we discussion the symplectic model reduction for Hamiltonian systems and introduce the Reduced Dissipative Hamiltonian method. Accuracy, stability and efficiency of the RDH method is then discussed in Section \ref{sec:4}, through simulation of the dissipative wave equation and a linear port-Hamiltonian system of an electrical circuit. We offer conclusive remarks in Section \ref{sec:5}.

### 2017-03-19 dissipative_hamiltonian/chapters/2.Hamil.tex
first draft finished

\section{Dissipative Hamiltonian Systems} \label{sec:2}
In this section we first discuss Hamiltonian systems. Add more later

\subsection{Hamiltonian Systems} Suppose that $(\mathbb{R}^{2n},\Omega)$ is a symplectic linear vector space, where $\mathbb{R}^{2n}$ is a configuration space and $\Omega:\mathbb{R}^{2n}\times \mathbb{R}^{2n} \to \mathbb R$ is a closed, skew-symmetric and non-degenerate 2-form on $\mathbb{R}^{2n}$. Given a smooth Hamiltonian function $H:\mathbb{R}^{2n}\to \mathbb R$, \emph{Hamilton}'s equations of evolution are given as
\begin{equation} \label{eq:2.1}
	\begin{aligned}
	\dot {z}(t) &= \mathbb J_{2n} \nabla_{z} H, \\
	z(0) &= z_0,
	\end{aligned}
\end{equation}
where $z \in\mathbb R^{2n}$ is the configuration coordinates and $\mathbb J_{2n}$ is a $2n\times 2n$ matrix such that $\Omega(x,y) = x^T \mathbb J_{2n} y$ \cite{Marsden:2010:IMS:1965128}. One by using the \emph{Symplectic Gram-Schmidt} \cite{de2006symplectic} can construct a coordinate system in which $\mathbb J_{2n}$ takes the form
\begin{equation} \label{eq:2.2}
	\mathbb{J}_{2n} = 
	\begin{pmatrix}
		0_n & I_n \\
		-I_n & 0_n
	\end{pmatrix},
\end{equation}
where $I_n$ and $0_n$ are the identity matrix and the zero matrix of size $n$ respectively. A main feature of Hamiltonian systems is the conservation of the Hamiltonian along the integral curves.
\begin{theorem} \label{theorem:2.1}
\cite{Marsden:2010:IMS:1965128} Consider the flow $\phi_t:\mathbb R\times \mathbb R^{2n} \to \mathbb R^{2n}$ of the Hamiltonian system (\ref{eq:2.1}). Then $H\circ \phi_t = H$.
\end{theorem}

In many physical problems $H$ represents the system energy and is bounded below. Here, we mostly assume that $H$ is a quadratic Hamiltonian, i.e., it takes the form $H(z) = \frac 1 2 z^T K^T K z$, where $K$ is a full rank $2n\times 2n$ matrix. This assumption leads to a linear system of evolution (\ref{eq:2.1}). Note that the Hamiltonian extension in section ?? and the model reduction for dissipative Hamiltonian systems in section ?? can be naturally extended to Hamiltonians of the form $H(z) = \frac 1 2 z^T K^T K z + g(z)$, where $g:\mathbb R^{2n} \to \mathbb R$ is an arbitrary function of $z$.

Under general coordinate changes, equations of evolution do not necessarily take the form of (\ref{eq:2.1}). Let $(\mathbb R^{2n},\Omega)$ and $(\mathbb R^{2k},\Lambda)$ be two symplectic linear vector spaces. A linear transformation $\alpha :\mathbb R^{2n} \to \mathbb R^{2k}$ is called a \emph{symplectic transformation} \cite{Marsden:2010:IMS:1965128} if
\begin{equation}
	\Omega(x,y) = \Lambda(\alpha(x),\alpha(y)), \quad \text{for all } x,y\in \mathbb R^{2n}.
\end{equation}
Alternatively in matrix notation, $A\in \mathbb R^{2n\times 2k}$ is called a \emph{symplectic matrix} if
\begin{equation}
	A^T \mathbb{J}_{2n} A = \mathbb{J}_{2k},
\end{equation}
where the superscript $T$ represents the transpose operator. As the symplectic 2-form is preserved under symplectic transformations, the form of the equations of evolution remains invariant \cite{Marsden:2010:IMS:1965128}. The \emph{symplectic inverse} of $A$, is a form of pseudo-inverse given by
\begin{equation}
	A^+ = \mathbb{\mathbb J}_{2k}^T A^T \mathbb J_{2n}.
\end{equation}
It is shown in \cite{Peng:2014di} that $A^+A = I_{2k}$ and that $(A^+)^T$ is a symplectic matrix. Further, one can easily check that $AA^+$ is idempotent and so is a projection operator onto the column span of $A$.

It is known that symplectic matrices are usually ill-conditioned \cite{Karow:2006cf}. Under some conditions on a symplectic space \cite{da2003introduction}, one can construct a symplectic basis which is also ortho-normal, and thus norm bounded. A basis which is both symplectic and orthogonal is referred to as an \emph{ortho-symplectic basis}. We refer the reader to \cite{da2003introduction} for conditions on existence and construction of an ortho-symplectic basis.

@@

\begin{equation} \label{eq:2.3}
\begin{aligned}
	p_{n+1/2} &= p_n - \frac{\Delta t}{2} \nabla_qH(q_{n},p_{n+1/2}), \\
	q_{n+1} &= q_n + \frac{\Delta t}{2} \left( \nabla_pH(q_{n},p_{n+1/2}) + \nabla_pH(q_{n+1},p_{n+1/2}) \right),\\
	p_{n+1} &= p_{n+1/2} - \frac{\Delta t}{2} \nabla_qH(q_{n+1},p_{n+1/2}),
\end{aligned}
\end{equation}
where $p$ and $q$ are the canonical coordinates. More information on construction and application of symplectic integrators can be found in \cite{Hairer:1250576}.

\subsection{Dissipative Hamiltonian Systems and Hamiltonian Extensions}

Many systems in engineering and physic appear as a perturbation of a Hamiltonian system, where the perturbation is regarded as dissipation. In these systems, the energy tends to decrease over time evolution, and the conservation law in Theorem \ref{theorem:2.1} does not hold anymore. Therefore, it is common to take the conservation of energy as a fundamental principle and consider the dissipative system coupled with a heat bath that absorbs the dissipated energy of the original system.

To account dissipation for a quadratic Hamiltonian $H(z) = \frac 1 2 z^T K^T K  z$, we rewrite (\ref{eq:2.1}) as a time dispersive and dissipative (TDD) \cite{Figotin:2006jy} system 
\begin{equation} \label{eq:2.4}
	\begin{aligned}
		& \dot {z} = \mathbb J_{2n} K^T f(t), \\
		& z(0) = z_0,
	\end{aligned}
\end{equation}
where $f$ is the solution to the integral equation
\begin{equation} \label{eq:2.4.1}
	f(t) + \int_0^t \chi(t-s) \cdot f(s)\ ds = K z.
\end{equation}
Here $\chi:\mathbb R^+\to \mathbb R^{2n\times 2n}$ is a bounded matrix valued function with respect to the Frobenius norm and is called the \emph{general susceptibility}. Note that integral term in (\ref{eq:2.4.1}) corresponds to the dissipation, whereas if $\chi(s) = 0$ then (\ref{eq:2.3}) is equivalent to (\ref{eq:2.1}). Note that under suitable assumptions on $K$, both the systems (\ref{eq:2.1}) and (\ref{eq:2.4}) are well-posed \cite{Figotin:2006jy}.

@@

It is shown in \cite{Figotin:2006jy,Figotin:2005} that under natural assumptions on a linear susceptibility $\chi(t)$ (see below), one can couple a TDD system of the form (\ref{eq:2.4}) with a canonical heat bath where the dissipated energy is captured in the heat bath in a canonical sense. In other words, one can construct a Hilbert space $\mathcal H$ and an isometric injection $I:\mathbb R^{2n} \to \mathbb R^{2n}\times \mathcal H^{2n}$ where the solution $z$ to (\ref{eq:2.4}) is the projection of $x$ onto $\mathbb R^{2n}$, where $x$ is the solution to
\begin{equation} \label{eq:2.7}
	\frac{d}{dt} x = J_{2n} \frac{\delta h}{\delta x}.
\end{equation}
Here $J_2n$ is the symplastic operator and $h:\mathbb R^{2n}\times \mathcal H^{2n} \to \mathbb R$ is an extended quadratic Hamiltonian function defined on $\mathbb R^{2n}\times \mathcal H^{2n}$.

\begin{theorem}
Suppose that $K$ is full ranked and $\chi(t)$ is symmetric. There exists a quadratic extension of the form (\ref{eq:2.8}) to (\ref{eq:2.4}), if
\begin{equation} \label{eq:2.8}
	\text{Im}(\xi\hat{\chi}(\xi)) \geq 0, \quad \forall \xi = \omega + i\eta, \ \eta \geq 0,
\end{equation}
where $\hat{\chi}$ is the Fourier-Laplace transform of $\chi$
\begin{equation} \label{eq:2.9}
	\hat{\chi}(\xi) = \int_0^\infty e^{i\xi t} \chi(t)\ dt.
\end{equation}
\end{theorem}

\begin{proof}
Here we prove the theorem for the case where $\chi$ is a constant symmetric matrix, where condition (\ref{eq:2.8}) corresponds to $\chi$ being positive semi-definite. We refer the reader to \cite{Figotin:2006jy} for the proof of the general case. Consider the Hamiltonian system
\begin{subequations}
\begin{align}
		\label{eq:2.10.a} & \dot{z}(t) = \mathbb J_{2n} K^T f(t), \\
		\label{eq:2.10.b} & \partial_t \theta(t,x) = \partial_x^2 \phi(t,x) + \sqrt 2 \delta_0(x) \cdot \sqrt{\chi}  f(t), \\
		\label{eq:2.10.c} & \partial_t \phi(t,x) = \theta(t,x),
\end{align}
\end{subequations}
together with the initial condition
\begin{equation} \label{eq:2.10.1}
	z(0) = z_0,\quad \theta(0,\cdot) = 0, \quad \phi(0,\cdot) = 0.
\end{equation}
Here $\theta$ and $\phi$ are vector valued functions in $\mathcal H^{2n}$, $\delta_0(s)$ is the Dirac's delta function, $\sqrt{ \chi}$ is the matrix square root of $\chi$ and $f$ is the solution to the equation
\begin{equation} \label{eq:2.11}
	f(t) + \sqrt{2} \cdot \sqrt{ \chi } \phi(t,0) = Kz(t).
\end{equation}
To show that the Hamiltonian system (\ref{eq:2.10.a}) to (\ref{eq:2.10.c}) is an extension to (\ref{eq:2.4}) in the sense discussed before, it is enough to show that the solution $f$ to the equation (\ref{eq:2.11}) also satisfies (\ref{eq:2.4.1}). Equations (\ref{eq:2.10.b}) and (\ref{eq:2.10.c}) are equations to a vibrating string, and can be solved analytically
\begin{equation} \label{eq:2.12}
	\phi(t,x) = \frac {\sqrt 2} 2 \int_0^{t-|x|} \sqrt{\chi} f(s)\ ds,\quad \theta(t,x) = \frac{\sqrt 2}{2} \cdot \sqrt{\chi} f(t - |x|).
\end{equation}
\end{proof}
We can recover (\ref{eq:2.4.1}) by substituting (\ref{eq:2.12}) into (\ref{eq:2.11}). The extended Hamiltonian $H_\text{ex}$ for the system (\ref{eq:2.10.a}) to (\ref{eq:2.10.c}) takes the quadratic from
\begin{equation} \label{eq:2.13}
	H_\text{ex}(z,\phi,\theta) = \frac 1 2 \left( \| Kz - \phi(t,0) \|_2^2 + \| \theta(t) \|^2_{\mathcal H^{2n} } + \| \partial_x\phi(t)\|^2_{\mathcal H^{2n} }\right)
\end{equation}
where $\| \cdot \|_2$ is the Euclidean norm on $\mathbb R^{2n}$ and $\| \cdot \|_{\mathcal H^{2n}}$ is the induced norm from the inner product on $\mathcal H^{2n}$.

Equations (\ref{eq:2.10.a}) and (\ref{eq:2.10.c}) are called the \emph{hidden strings}. The dissipation of energy in the original system (\ref{eq:2.4}) is carried away, as vibrations, along the added strings which makes the extended system conservative. Hamiltonian extension of the damped harmonic oscillator in Example \ref{example:2.1} is exactly the Lamb model \cite{lamb:1900}. The Lamb model is a harmonic oscillator coupled with a vibrating string, and the tension in the string causes linear dissipation in the dynamics of the harmonic oscillator.

Note that time integration of (\ref{eq:2.10.a})-(\ref{eq:2.10.c}) involves the integration of $f$ in (\ref{eq:2.12}). In general $f(t)$ must be stored and may cause serious storage limitation in long-time integration. However, we are interested solely in finding $z(t)$ which depends on $f(t)$, and $\phi(t,0)$ that is the integral of the history of $f(t)$. So by carefully choosing a quadrature rule that uses the same quadrature nodes as the time integrator we can avoid storing the history of $f$. For example for the trapezoidal rule, we recover the recursive relation
\begin{equation} \label{eq:2.14}
	\int_{0}^{t_n} f(s) \ ds \approx \frac{\Delta t}{2} f(t_n) + \frac{\Delta t}{2} f(t_{n-1}) + \int_{0}^{t_{n-1}} f(s) \ ds,
\end{equation}
where $\Delta t$ is the time discretization step. The recursive relation in (\ref{eq:2.12}) suggests that only storing the value of the integral term together with the state of $f$ in the previous time step is required to evaluate the integral for the new time step. For other interpolation based quadrature rules, we can construct similar recursive rules of the form
\begin{equation}
	\int_{0}^{t_n} f(s) \ ds \approx \sum_{i=0}^{k} \omega_i f(t_{n-i})  + \int_{0}^{t_{n-k}} f(s) \ ds
\end{equation}
for some quadrature weights $\omega_i$, $i=1,\dots,k$ with $k\ll n$. Thus, time integration of (\ref{eq:2.10.a})-(\ref{eq:2.10.c}), requires storing only $k$ evaluations of $f$.

>>>>

\section{Dissipative Hamiltonian Systems} \label{sec:2}
In this section, we first summarize basic concepts in the geometry of symplectic linear vector space. Then, we discuss Hamiltonian systems under dissipative forces and introduce the Hamiltonian extension of dissipative Hamiltonian systems.

\subsection{Hamiltonian Systems} \label{sec:2.1} Suppose that $(\mathbb{R}^{2n},\Omega)$ is a symplectic linear vector space, where $\mathbb{R}^{2n}$ is a configuration space and $\Omega:\mathbb{R}^{2n}\times \mathbb{R}^{2n} \to \mathbb R$ is a closed, skew-symmetric and non-degenerate 2-form on $\mathbb{R}^{2n}$. Given a smooth Hamiltonian function $H:\mathbb{R}^{2n}\to \mathbb R$, \emph{Hamilton}'s equations of evolution are given as
\begin{equation} \label{eq:2.1}
	\begin{aligned}
	\dot {z}(t) &= \mathbb J_{2n} \nabla_{z} H, \\
	z(0) &= z_0,
	\end{aligned}
\end{equation}
where $z \in\mathbb R^{2n}$ is the configuration coordinates and $\mathbb J_{2n}$ is a $2n\times 2n$ matrix such that $\Omega(x,y) = x^T \mathbb J_{2n} y$, for all $x,y\in \mathbb R^{2n}$ \cite{Marsden:2010:IMS:1965128}. One by using the \emph{Symplectic Gram-Schmidt} \cite{de2006symplectic} can construct a coordinate system in which $\mathbb J_{2n}$ takes the form
\begin{equation} \label{eq:2.2}
	\mathbb{J}_{2n} = 
	\begin{pmatrix}
		0_n & I_n \\
		-I_n & 0_n
	\end{pmatrix},
\end{equation}
where $I_n$ and $0_n$ are the identity matrix and the zero matrix of size $n$ respectively. A main feature of Hamiltonian systems is the conservation of the Hamiltonian along the integral curves.
\begin{theorem} \label{theorem:2.1}
\cite{Marsden:2010:IMS:1965128} Consider the flow $\phi_t:\mathbb R\times \mathbb R^{2n} \to \mathbb R^{2n}$ of the Hamiltonian system (\ref{eq:2.1}). Then $H\circ \phi_t = H$.
\end{theorem}

In many physical problems, $H$ represents the system energy and is bounded below. Here, to avoid difficulties with well-posedness of Hamiltonian systems,  we mostly assume that $H$ is a quadratic Hamiltonian, i.e., it takes the form $H(z) = \frac 1 2 z^T K^T K z$, where $K$ is a full rank $2n\times 2n$ matrix. This assumption leads to a linear system of evolution (\ref{eq:2.1}). Note that the Hamiltonian extension in section \ref{sec:2.2} and the Reduced Dissipative Hamiltonian method in section \ref{sec:3.3} can be naturally extended to Hamiltonians of the form $H(z) = \frac 1 2 z^T K^T K z + g(z)$, where $g:\mathbb R^{2n} \to \mathbb R$ is an arbitrary function of $z$.

Under general coordinate transformations, the equations of evolution do not necessarily take the form of (\ref{eq:2.1}). Let $(\mathbb R^{2n},\Omega)$ and $(\mathbb R^{2k},\Lambda)$ be two symplectic linear vector spaces. A linear transformation $\alpha :\mathbb R^{2n} \to \mathbb R^{2k}$ is called a \emph{symplectic transformation} \cite{Marsden:2010:IMS:1965128} if
\begin{equation}
	\Omega(x,y) = \Lambda(\alpha(x),\alpha(y)), \quad \text{for all } x,y\in \mathbb R^{2n}.
\end{equation}
Alternatively in matrix notation, $A\in \mathbb R^{2n\times 2k}$ is called a \emph{symplectic matrix} if
\begin{equation}
	A^T \mathbb{J}_{2n} A = \mathbb{J}_{2k},
\end{equation}
where the superscript $T$ represents the transpose operator. As the symplectic 2-form is preserved under a symplectic transformation, the form of the equations of evolution remains invariant through a symplectic coordinate change  \cite{Marsden:2010:IMS:1965128}. The \emph{symplectic inverse} of $A$, is a pseudo-inverse given by
\begin{equation}
	A^+ = \mathbb{\mathbb J}_{2k}^T A^T \mathbb J_{2n}.
\end{equation}
It is shown in \cite{Peng:2014di} that $A^+A = I_{2k}$ and that $(A^+)^T$ are symplectic matrices. Further, one can easily check that $AA^+$ is idempotent, meaning it is a projection operator onto the column span of $A$.

It is known that symplectic matrices are usually ill-conditioned \cite{Karow:2006cf}. Under some conditions on a symplectic space \cite{da2003introduction}, one can construct a symplectic basis which is also ortho-normal, and thus norm bounded. A basis which is both symplectic and orthogonal is called a \emph{ortho-symplectic basis}. We refer the reader to \cite{da2003introduction} for conditions on existence and construction of an ortho-symplectic basis.

@@

\begin{equation} \label{eq:2.3}
\begin{aligned}
	p_{n+1/2} &= p_n - \frac{\Delta t}{2} \nabla_qH(q_{n},p_{n+1/2}), \\
	q_{n+1} &= q_n + \frac{\Delta t}{2} \left( \nabla_pH(q_{n},p_{n+1/2}) + \nabla_pH(q_{n+1},p_{n+1/2}) \right),\\
	p_{n+1} &= p_{n+1/2} - \frac{\Delta t}{2} \nabla_qH(q_{n+1},p_{n+1/2}),
\end{aligned}
\end{equation}
where $p$ and $q$ are the canonical coordinates $z = (q^T,p^T)^T$. More information on construction and application of symplectic integrators can be found in \cite{Hairer:1250576}.

\subsection{Dissipative Hamiltonian Systems and Hamiltonian Extensions} \label{sec:2.2}

Many systems in engineering and physic appear as a perturbation of a Hamiltonian system, where the perturbation is regarded as dissipation. In these systems, the energy tends to decrease over time evolution, and thus, the conservation law in Theorem \ref{theorem:2.1} does not hold anymore. Therefore, it is common to take the conservation of energy as a fundamental principle and consider the dissipative system coupled with a heat bath that absorbs the dissipated energy of the original system.

To account dissipation for a quadratic Hamiltonian $H(z) = \frac 1 2 z^T K^T K  z$, we rewrite (\ref{eq:2.1}) as a time dispersive and dissipative (TDD) \cite{Figotin:2006jy} system 
\begin{equation} \label{eq:2.4}
	\begin{aligned}
		& \dot {z} = \mathbb J_{2n} K^T f(t), \\
		& z(0) = z_0,
	\end{aligned}
\end{equation}
where $f$ is the solution to the integral equation
\begin{equation} \label{eq:2.4.1}
	f(t) + \int_0^t \chi(t-s) \cdot f(s)\ ds = K z.
\end{equation}
Here $\chi:\mathbb R^+\to \mathbb R^{2n\times 2n}$ is a bounded matrix valued function with respect to the Frobenius norm and is called the \emph{general susceptibility}. Note that the integral term in (\ref{eq:2.4.1}) corresponds to the dissipation, whereas if $\chi(s) = 0$ then (\ref{eq:2.4}) is equivalent to (\ref{eq:2.1}). Furthermore, under suitable assumptions on $K$, both the systems (\ref{eq:2.1}) and (\ref{eq:2.4}) are well-posed \cite{Figotin:2006jy}.

@@

It is shown in \cite{Figotin:2006jy,Figotin:2005} that under natural assumptions on the linear susceptibility $\chi(t)$ (see below), one can couple a TDD system of the form (\ref{eq:2.4}) with a canonical heat bath where the dissipated energy is captured in the heat bath in a canonical sense. In other words, one can construct a Hilbert space $\mathcal H$ and an isometric injection $I:\mathbb R^{2n} \to \mathbb R^{2n}\times \mathcal H^{2n}$ where the solution $z$ to (\ref{eq:2.4}) is the projection of $x$ onto $\mathbb R^{2n}$, where $x$ is the solution to
\begin{equation} \label{eq:2.7}
	\frac{d}{dt} x = \mathcal J_{2n} \frac{\delta H_{\text{ex}}}{\delta x}.
\end{equation}
Here $H_{\text{ex}}:\mathbb R^{2n}\times \mathcal H^{2n} \to \mathbb R$ is an extended quadratic Hamiltonian function and $\mathcal J_{2n}$ is the symplastic operator defined on $\mathbb R^{2n}\times \mathcal H^{2n}$ respectively.

\begin{theorem}
Suppose that $K$ is full ranked and $\chi(t)$ is symmetric. There exists a quadratic extension of the form (\ref{eq:2.7}) to (\ref{eq:2.4}), if
\begin{equation} \label{eq:2.8}
	\text{Im}(\xi\hat{\chi}(\xi)) \geq 0, \quad \forall \xi = \omega + i\eta, \ \eta \geq 0,
\end{equation}
where $\hat{\chi}$ is the Fourier-Laplace transform of $\chi$
\begin{equation} \label{eq:2.9}
	\hat{\chi}(\xi) = \int_0^\infty e^{i\xi t} \chi(t)\ dt.
\end{equation}
\end{theorem}

\begin{proof}
Here we prove the theorem for the case where $\chi$ is a constant symmetric matrix, where condition (\ref{eq:2.8}) corresponds to $\chi$ being positive semi-definite. We refer the reader to \cite{Figotin:2006jy} for the proof of the general case. Consider the Hamiltonian system
\begin{subequations}
\begin{align}
		\label{eq:2.10.a} & \dot{z}(t) = \mathbb J_{2n} K^T f(t), \\
		\label{eq:2.10.b} & \partial_t \theta(t,x) = \partial_x^2 \phi(t,x) + \sqrt 2 \delta_0(x) \cdot \sqrt{\chi}  f(t), \\
		\label{eq:2.10.c} & \partial_t \phi(t,x) = \theta(t,x),
\end{align}
\end{subequations}
together with the initial condition
\begin{equation} \label{eq:2.10.1}
	z(0) = z_0,\quad \theta(0,\cdot) = 0, \quad \phi(0,\cdot) = 0.
\end{equation}
Here $\theta$ and $\phi$ are vector valued functions in $\mathcal H^{2n}$, $\delta_0(s)$ is the Dirac's delta function, $\sqrt{ \chi}$ is the matrix square root of $\chi$ and $f$ is the solution to the equation
\begin{equation} \label{eq:2.11}
	f(t) + \sqrt{2} \cdot \sqrt{ \chi } \phi(t,0) = Kz(t).
\end{equation}
To show that the Hamiltonian system (\ref{eq:2.10.a}-\ref{eq:2.10.c}) is an extension to (\ref{eq:2.4}) in the sense discussed above, it is enough to show that the solution $f$ to the equation (\ref{eq:2.11}) also satisfies (\ref{eq:2.4.1}). Equations (\ref{eq:2.10.b}-\ref{eq:2.10.c}) are equations to a vibrating string, and can be solved analytically
\begin{equation} \label{eq:2.12}
	\phi(t,x) = \frac {\sqrt 2} 2 \int_0^{t-|x|} \sqrt{\chi} f(s)\ ds,\quad \theta(t,x) = \frac{\sqrt 2}{2} \cdot \sqrt{\chi} f(t - |x|).
\end{equation}
\end{proof}
Then, we can recover (\ref{eq:2.4.1}) by substituting (\ref{eq:2.12}) into (\ref{eq:2.11}). The extended Hamiltonian $H_\text{ex}$ for the system (\ref{eq:2.10.a}-\ref{eq:2.10.c}) takes the quadratic from
\begin{equation} \label{eq:2.13}
	H_\text{ex}(z,\phi,\theta) = \frac 1 2 \left( \| Kz - \phi(t,0) \|_2^2 + \| \theta(t) \|^2_{\mathcal H^{2n} } + \| \partial_x\phi(t)\|^2_{\mathcal H^{2n} }\right)
\end{equation}
where $\| \cdot \|_2$ is the Euclidean norm on $\mathbb R^{2n}$ and $\| \cdot \|_{\mathcal H^{2n}}$ is the induced norm from the inner product on $\mathcal H^{2n}$.

Equations (\ref{eq:2.10.b}) and (\ref{eq:2.10.c}) are called the \emph{hidden strings}. The dissipation of energy in the original system (\ref{eq:2.4}) is carried away, as vibrations, along the added strings which makes the extended system conservative. The Hamiltonian extension of the damped harmonic oscillator in Example \ref{example:2.1} is exactly the Lamb model \cite{lamb:1900}. The Lamb model is a harmonic oscillator coupled with a vibrating string, and the tension in the string causes linear dissipation in the dynamics of the harmonic oscillator.

Note that the time integration of (\ref{eq:2.10.a}-\ref{eq:2.10.c}) involves the integration of $f$ in (\ref{eq:2.12}). In general, the history of $f(t)$ must be stored and may cause serious storage limitation in long-time integration. However, we are interested solely in finding $z(t)$ which depends on $f$ at time $t$, and $\phi(t,0)$, that is the integral of the history of $f$. So by carefully choosing a quadrature rule that uses the same quadrature nodes as the time integrator we can avoid storing the history of $f$. For example for the trapezoidal rule, we recover the recursive relation
\begin{equation} \label{eq:2.14}
	\int_{0}^{t_n} f(s) \ ds \approx \frac{\Delta t}{2} f(t_n) + \frac{\Delta t}{2} f(t_{n-1}) + \int_{0}^{t_{n-1}} f(s) \ ds,
\end{equation}
where $\Delta t$ is the time discretization step. The recursive relation in (\ref{eq:2.12}) suggests that only storing the value of the integral term together with the state of $f$ in the previous time step is required to evaluate the integral for the new time step. For other interpolation based quadrature rules, we can construct similar recursive rules of the form
\begin{equation}
	\int_{0}^{t_n} f(s) \ ds \approx \sum_{i=0}^{k} \omega_i f(t_{n-i})  + \int_{0}^{t_{n-k}} f(s) \ ds
\end{equation}
for some quadrature weights $\omega_i$, $i=1,\dots,k$ with $k\ll n$. Thus, time integration of (\ref{eq:2.10.a}-\ref{eq:2.10.c}), requires storing only $k$ evaluations of $f$.
